Can Cultural Transmission Explain The Evolution of Cooperative Behavior?

### 2020-04-11 ms/main.tex

Therefore, the frequency $p'$ of phenotype $A$ among adults in the next generation, after horizontal transmission, is 
\begin{equation}\label{eq:nextgen_adults}
\begin{aligned}
p'
& = \hat{p}^2 [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A]
\end{aligned}
\end{equation}

The frequency of $A$ among parents follows a similar dynamic, but also includes the effect of natural selection, and is therefore
\begin{equation}\label{eq:nextgen_parents}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) (1-c) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} (1+b) [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A],
\end{aligned}
\end{equation}
where $\bar{w} = \hat{p} w_A + (1-\hat{p}) w_B = 1 + \hat{p}(b-c)$ is the population mean fitness.

@@

With only vertical transmission eq.~\ref{eq:horizontal} becomes
$\hat{p} =  \tilde{p}$,
and eq.~\ref{eq:nextgen_parents} for the change in frequency $p'$ of phenotype $A$ among parents can be written as

\begin{equation} 
\begin{split}\label{eq:nextgen_parents_vertical_only}
\bar{w} \tilde{p}' 
& = \tilde{p}^2 (1+b-c) (1 - (1-\tilde{p}) (1-\alpha) T_B) \\
& + \tilde{p}(1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
& + \tilde{p}(1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
& + (1-\tilde{p})^2 \tilde{p} (1-\alpha) T_A
\end{split}
\end{equation}

\textbf{Result 1: Vertical and horizontal transmission of cooperation.}
If 
\begin{equation}\label{eq:unequal_transmission}
c \cdot (1-T_B) < b \cdot  \alpha T_A  + (T_A - T_B) [1 + bp(1-\alpha)]
\end{equation}
then $\tilde{p}' > \tilde{p}$, and the frequency of phenotype $A$ among parents increases. 
\\

@@

In this case horizontal transmission can not occur as a result of interaction.  
Then, eq.~\ref{eq:unequal_transmission} becomes:
\begin{equation} 
\begin{split} \label{eq:horizontal_transmission_from_population}
c\cdot(1-T_B) < & (T_A-T_B)(1+bp)  
\end{split}
\end{equation}

Therefore, since $c>0$, cooperation will not take over the population (and furthermore will become extinct) if $T_A\leq T_B$.
When $T_A > T_B$ eq.~\ref{eq:horizontal_transmission_from_population} becomes:
\begin{equation} 
\begin{split} \label{eq:horizontal_transmission_from_population_reorder}
p & > \frac{c\cdot(1-T_B)+T_B-T_A}{b\cdot(T_A-T_B)} .
\end{split}
\end{equation}
Hence, cooperation will take over the population when when eq.~\ref{eq:horizontal_transmission_from_population_reorder} is satisfied, that is if $p$ is above the threshold on the RHS. 
Specifically, a sufficient condition for cooperation fixation is found by setting the RHS to be negative:
\begin{equation} 
  \begin{split} \label{eq:sufficient_horizontal_transmission_from_population}
c < &\, \frac{T_A-T_B}{1-T_B}  
  \end{split}
  \end{equation}

>>>>

Therefore, the frequency $p'$ of phenotype $A$ among adults in the next generation, after horizontal transmission, is 
\begin{equation}\label{eq:nextgen_adults}
\begin{aligned}
p'
& = \hat{p}^2 [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A]
\end{aligned}
\end{equation}
Which can be simplified into
\begin{equation}\label{eq:nextgen_adults_slimpify}
  \begin{aligned}
  p'  & = \hat{p}^2(T_B-T_A) + \hat{p}(1+T_A-T_B) \\
  \end{aligned}
  \end{equation}

The frequency of $A$ among parents follows a similar dynamic, but also includes the effect of natural selection, and is therefore
\begin{equation}\label{eq:nextgen_parents}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) (1-c) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} (1+b) [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A],
\end{aligned}
\end{equation}
where $\bar{w} = \hat{p} w_A + (1-\hat{p}) w_B = 1 + \hat{p}(b-c)$ is the population mean fitness.
Eq~\ref{eq:nextgen_parents} can be slimpified to
\begin{equation}\label{eq:nextgen_parents_simplified}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) [1-(1-\hat{p}(1-\alpha)T_B)] \\
& + \hat{p}(1-\hat{p}) (1-c) [\hat{p}(1-\alpha)T_B+1-T_B] \\
& + (1-\hat{p})\hat{p} (1+b) [\hat{p}(1-\alpha) + \alpha]T_A \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p}]T_A,
\end{aligned}
\end{equation}

@@

With only vertical transmission eq.~\ref{eq:horizontal} becomes
$\hat{p} =  \tilde{p}$,
and eq.~\ref{eq:nextgen_parents_simplified} for the change in frequency $p'$ of phenotype $A$ among parents can be written as

\begin{equation} 
\begin{split}\label{eq:nextgen_parents_vertical_only}
\bar{w} \tilde{p}' 
& = \tilde{p}^2 (1+b-c) [1 - (1-\tilde{p}) (1-\alpha) T_B] \\
& + \tilde{p}(1-\tilde{p}) (1-c) [\tilde{p} (1-\alpha) T_B + 1 - T_B] \\
& + \tilde{p}(1-\tilde{p}) (1+b) [\tilde{p} (1-\alpha) + \alpha] T_A \\
& + (1-\tilde{p})^2 \tilde{p} (1-\alpha) T_A
\end{split}
\end{equation}

\textbf{Result 1: Vertical and horizontal transmission of cooperation.}
If 
\begin{equation} \label{eq:unequal_transmission}
c \cdot (1-T_B) < b \cdot  \alpha T_A  + (T_A - T_B) [1 + b\tilde{p}(1-\alpha)]
\end{equation}
then $\tilde{p}' > \tilde{p}$, and the frequency of phenotype $A$ among parents increases. 
\\

@@

In this case horizontal transmission can not occur as a result of interaction.  
Then, eq.~\ref{eq:unequal_transmission} becomes:
\begin{equation} 
\begin{split} \label{eq:horizontal_transmission_from_population}
c\cdot(1-T_B) < & (T_A-T_B)(1+b\tilde{p})  
\end{split}
\end{equation}

Therefore, since $c>0$, cooperation will not take over the population (and furthermore will become extinct) if $T_A\leq T_B$.
When $T_A > T_B$ eq.~\ref{eq:horizontal_transmission_from_population} becomes:
\begin{equation} 
\begin{split} \label{eq:horizontal_transmission_from_population_reorder}
\tilde{p} & > \frac{c\cdot(1-T_B)+T_B-T_A}{b\cdot(T_A-T_B)}
\end{split}
\end{equation}
Hence, cooperation will take over the population when when eq.~\ref{eq:horizontal_transmission_from_population_reorder} is satisfied, that is if $\tilde{p}$ is above the threshold on the RHS. 
When $\tilde{p}$ is equal to eq.~\ref{eq:horizontal_transmission_from_population_reorder} RSH the frequency of phenotype A do not change. Denote $\tilde{p}^{*}$:
\begin{equation} 
\begin{split} \label{eq:horizontal_transmission_from_population_equilibrium}
\tilde{p}^{*} & = \frac{c\cdot(1-T_B)+T_B-T_A}{b\cdot(T_A-T_B)} .
\end{split}
\end{equation}
Therefore, $\tilde{p}^{*}$ is the third equilibrium in this system.
Specifically, a sufficient condition for cooperation fixation is found by setting the the equilibrium $\tilde{p}^*$  to be negative:
\begin{equation} 
\begin{split} \label{eq:sufficient_horizontal_transmission_from_population}
c < &\, \frac{T_A-T_B}{1-T_B}  
\end{split}
\end{equation}
